Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-27 paper/paper.tex

\begin{figure}[h]
\[
\begin{array}{lcll}
% \mathit{Natural\ numbers}\\
% [[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{conditional expression} \\ \\
% \mathit{Renaming}\\
% [[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitutions}\\
[[rho]] & \in & [[SNat -> STm]] & \\ \\
\mathit{Typing Contexts}\\
 [[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. The type $[[Set i]]$ represent universe types and $[[var n]]$
represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
$[[n]]$ are all metavariables representing natural numbers, we always use
$[[i]],[[j]]$ to represent universe levels and $[[n]]$ to represent term
variables.  We use de Bruijn indices for both our Coq development and the
informal presentation here because it is more amenable to automation
(Section~\ref{sec:automation}) and leaves no ambiguity about variable freeness
conditions. As a result, variables do not appear in binding forms, such as in
abstractions $[[ \ a ]]$ and dependent function types $[[Pi A B]]$.  This latter
form binds a new variable of type $[[A]]$ that can appear in the codomain of
the type $[[B]]$.  We use the notation $[[A -> B]]$ when the output type
$[[B]]$ is not dependent on the input variable.
We include in \lang{} the intensional identity type
$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
$[[J c a b p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
$[[b]]$, and $[[c]]$ is the term whose type is to be casted.
Finally, \lang{} includes booleans, with standard syntax.
\scw{Example of
 what you can use identity types for? Or an example of a program that uses
J?}

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
        [[idtm n]] & := & [[var n]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Weakening substitution: }} \uparrow^{n}} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]} \\ \\
        %
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        %
        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho .: a]]) } \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \multicolumn{3}{l}{\mbox{\emph{Lifting operation: }} \Uparrow\rho} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

The \lang language is expressive enough to support large
eliminations, the ability to compute a type using a term as input. For
example, the function $[[\ if var 0 Bool Bool -> Bool]]$ returns
either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
is $[[true]]$ or $[[false]]$.

We adapt from \citet{autosubst2} notation for simultaneous renaming,
substitution, and other auxiliary definitions used in our paper, summarized in
Figure~\ref{fig:auxdef}. We need these definitions to specify the reduction
and typing relations for \lang{}. The metavariable $[[rho]]$ represents
substitutions, which are mappings from de Bruijn indices to terms.  One
example of a substitution is the function $\ottkw{id}_{tm}$, the identity
substitution that maps each index to the corresponding term variable.  The
$\uparrow^{[[n]]}$ substitution is more general and adds $[[n]]$ to the index
before injecting it as a term. The extension operation, $[[(rho .: a)]]$,
allows us to extend a substitution $[[rho]]$ with an extra element $[[a]]$;
the result maps the variable $[[var 0]]$ to $[[a]]$ and variables
$[[var Suc i]]$ to $[[rho i]]$.

The substitution operator, which takes the form $[[a {rho}]]$,
traverses the syntax of $[[a]]$ and replaces each variable $[[var i]]$
with the term $[[rho i]]$. When traversing under binders (e.g. in the
$[[(\ a) { rho }]]$ case), the $\Uparrow$ operator prevents $[[var
0]]$, the bound variable, from being replaced by
$[[rho]]$. From the definition of $[[up rho]]$, we see that it keeps
$[[var 0]]$ unchanged during substitution and replaces each $[[var
Suc i]]$ with $[[rho i {up 1}]]$. The shifting with $[[up 1]]$
is required so the variables in terms from the codomain of $[[rho]]$ skip over
the newly introduced binder to refer to the correct binding
location.

The substitution operator is referred to as simultaneous substitution as it
substitutes for all variables at once. It is possible to recover single
substitution by composing the extension operator and the identity
substitution: $[[a { b }]] := [[a { idtm .: b }]]$. In our proofs, we find it
more convenient to formulate simultaneous substitution directly rather than
recovering it from single substitution.  In particular, this shows up in the
definition of semantic typing in Section~\ref{sec:logrelproof}, which relies
on simultaneous substitution.

@@

\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
% From Lemmas~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
% \ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
% equivalence relation.

@@

In \rref{T-Var}, $[[G n]]$ is the partial function defined by the equations
$[[ (G ++ A) 0 ]] := [[A]]$ and $[[ (G ++ A) Suc n ]] := [[G n]]$.  The
precondition $[[n < | G | ]]$, where $[[ | G | ]]$ represents the length of
the context $[[G]]$, ensures that $[[G n]]$ has a defined output. The type
$[[G n]]$ needs to be shifted by $[[Suc n]]$ since the types stored in the
context $[[G]]$ are scoped differently. In the typing judgment
$[[G |- a : A]]$, we want both $[[a]]$ and $[[A]]$ to have the same
scope. However, the closer a type is to the head of $[[G]]$, the smaller de
Bruijn it has since there is less variables in the context it can refer to
(see \rref{Ctx-Cons}). Shifting $[[G n]]$ by $[[Suc n]]$ weakens the type so
it is scoped consistently with the full context $[[G]]$ rather than a
truncated version of $[[G]]$.

Although the use of de Bruijn indices complicates the variable rule, the rules
that involve binding, such as \rref{T-Pi,T-Abs}, are straightforward.

@@

The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
matching.  The result type of this expression $[[A {a}]]$ is composed of some
\emph{motive} $[[A]]$, a type where its single free variable has been replaced
with the condition of the if expression. When typing the true branch, this
substitution replaces the variable by $[[true]]$, and similarly for the false
branch. As a result, the type system communicates the information gained from
the % run-time \yl{commented out run-time per Jonathan's comment}
test to each of the branches of the expression.

@@

% , the
% statement that $\Gamma \vdash [[Pi A0 B0]] \equiv [[Pi A1 B1]]$
% implies $\Gamma \vdash [[A0]] \equiv [[A1]]$ and
% $\Gamma, [[A0]]\vdash [[B0]] \equiv [[B1]]$.

@@

A naive attempt to proving Theorem~\ref{theorem:consistency} by
induction on the derivation $[[empty |- a : Void]]$ would succeed at
almost all cases except for \rref{T-App}. In the application
case, we are given $[[empty |- b : Pi A B]]$ and $[[empty |- a : A]]$, and
the equality that $[[B {a} = Bool]]$. Our goal is to show that
$[[empty |- b a : Void]]$ is not possible. However, note that there is
nothing we know of $[[b]]$ or $[[a]]$ from the induction hypothesis
because neither $[[Pi A B]]$ nor $[[A]]$ is equal to $[[Bool]]$.
We have no way of deriving a contradiction from $[[empty |- b a :
Void]]$. The takeaway from this failed attempt is that, in order to
derive the consistency, we need to know something about types other
than $[[Void]]$. From a pragmatic point of view, proof by logical
relation can be seen as a sophisticated way of strengthening the
induction hypothesis. From the strengthened property, the fundamental
theorem, we will be able to derive consistency as a corollary.

The complexity of applying proof by logical relation to dependent types stems
from the fact that the logical relation is much harder to define. In
simply typed languages, the logical relation is defined as a recursive
function over the type $[[A]]$.
In dependent types, the type
$[[A]]$ can take the form $[[(\ var 0) Bool]]$. To assign meaning to
this type, we need to first reduce it to $[[Bool]]$. However, we
cannot write a function that performs the reduction because we do not
know the termination of well-typed terms a priori. As a result, we
define the logical relation as an inductively defined relation,
reminiscent of how we specify the reduction graph of a partial
function; the functionality of the relation can later be recovered in Lemma~\ref{lemma:logreldeter}.

@@

% To derive consistency, it suffices to restrict $[[S]]$ to the set of
% terms that reduce to closed terms.
\Rref{I-Void, I-Bool} capture terms that \emph{behave like} the inhabitants
of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For
example, the $[[Void]]$ type should not have any inhabitants under the
empty context, where as the $[[Bool]]$ type corresponds to terms that
evaluate to $[[true]]$ or $[[false]]$. However, the characterization
of $[[Bool]]$ in our logical relation does not always correspond to
well-typed or even closed terms. For example, we have $[[if false var 0
true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
  true \/ a =>+ false }]]$ since it evaluates to $[[true]]$.
The independence of syntactic typing in our logical relation allows
our semantic typing definition in Section~\ref{sec:logrelproof} to be
meaningful on its own. Furthermore, not having to embed scoping
information into the logical relation avoids extra bookkeeping and the
need for a Kripke-style logical relation.
\yl{Not sure what to cite from Derek Dreyer. I know his blog post
  about semantic type soundness but is there a good paper to cite? one
of the rust papers?}

% unsurprising; when considering only closed terms, the empty type
% should not be inhabited and therefore corresponds to the empty set,
% whereas the boolean type is semantically inhabited by terms that
% evaluate to the boolean values $[[true]]$ or $[[false]]$.
\Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresponds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Red} enables us to reduce our types in order
to assign meanings. Recall the type expression $[[(\ var 0)
Bool]]$. \Rref{I-Red} says to know that
$[[Interp I i  (\ var 0) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\var 0) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ var 0) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Red}
and \rref{I-Bool}.

\Rref{I-Pi} is the most complicated rule in our logical
relation. Instead of explaining it directly, we consider the following
simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}[I-PiAlt admissibility]
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a } S0 }]]$.
\end{proof}
Compared to the \rref{I-Pi}, which contains the mysterious relation
$[[R]]$, \rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The precondition of
\rref{I-PiAlt} says that
the function type $[[Pi A B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a}]]$, obtained by substituting $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation. In its conclusion, we say
that a term $[[b]]$ belongs to an interpretation of $[[Pi A B]]$ if
for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
application form $[[b a]]$ belongs to all possible interpretations of
$[[B {a}]]$ (the pre-condition ensures at least one interpretation
exists for each $[[B {a}]]$ where $[[a in S]]$).

In fact, while \rref{I-PiAlt} is an instantiation of \rref{I-Pi},
these two rules are equivalent in the sense that every derivation of
involving \rref{I-Pi} can be systematically replaced by
\rref{I-PiAlt}. This equivalence follows directly from the fact
that the logical relation is a partial function, a result we will show
in Lemma~\ref{lemma:logreldeter}. The preconditions of \rref{I-Pi},
when combined with the functionality of the logical relation, uniquely
determine the relation $[[R in S * PowerSet STm]]$ to be the relation
$[[{ (a , S0 ) | a in S , Interp I  i B { a } S0 }]]$. This result is
formally shown through the improved inversion lemma for function types
(Lemma~\ref{lemma:piinvalt}).

Unfortunately, we cannot define the function case of our logical
relation directly using \rref{I-PiAlt} since the occurrence of
$[[Interp I i B {a} S0]]$ in its conclusion not only violates the
syntactic strict positivity constraint required in proof assistants,
but is genuinely non-monotone when we rewrite the inductive definition
as an endofunction over the domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition as the relation
grows, whereas the
relation $[[R]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ and
prevents the set of witnesses from growing. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties to justify the well-definedness of our
inductive relation with \rref{I-PiAlt}, we opt for our current
formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[R]]$ each time we apply it, though this is mitigated by
the immediate admissibility of \rref{I-PiAlt} and the alternative $\Pi$ inversion
principle (Lemma~\ref{lemma:piinvalt}), which we derive after proving functionality.

>>>>

\begin{figure}[h]
\[
\begin{array}{lcll}
% \mathit{Natural\ numbers}\\
% [[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[x]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi x : A . B]]\ |\ [[\ x . a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{conditional expression} \\ \\
% \mathit{Renaming}\\
% [[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitutions}\\
[[rho]] & \in & [[SVar -> STm]] & \\ \\
\mathit{Typing Contexts}\\
 [[G]]       & ::= & [[empty]]\ |\ [[G ++ x : A]] &  \\ \\
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. The type $[[Set i]]$ represent universe types.%  and $[[x]]$
% represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
% $[[n]]$ are all metavariables representing natural numbers, we always use
% $[[i]],[[j]]$ to represent universe levels and $[[n]]$ to represent term
% variables.
% We use de Bruijn indices for our Coq development, but use nominal
% representation in the paper because it is more amenable to automation
% (Section~\ref{sec:automation}) and leaves no ambiguity about variable freeness
% conditions.
\yl{need to fix the text}
As a result, variables do not appear in binding forms, such as in
abstractions $[[ \ x . a ]]$ and dependent function types $[[Pi x : A . B]]$.  This latter
form binds a new variable of type $[[A]]$ that can appear in the codomain of
the type $[[B]]$.  We use the notation $[[A -> B]]$ when the output type
$[[B]]$ is not dependent on the input variable.
We include in \lang{} the intensional identity type
$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
$[[J c a b p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
$[[b]]$, and $[[c]]$ is the term whose type is to be casted.
Finally, \lang{} includes booleans, with standard syntax.
\scw{Example of
 what you can use identity types for? Or an example of a program that uses
J?}

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} [[idtm]]} \\
        [[idtm x]] & := & [[x]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho
        .: x -> a]]) } \\
        ([[rho .: x -> a]])([[x]]) & := & [[a]] \\
        ([[rho .: x -> a]])([[y]]) & := & [[rho y]] \text { if  } y
                                          \neq x\\ \\
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \multicolumn{3}{l}{\mbox{\emph{Substitution operation: }} ([[ a { rho } ]])} \\
        [[x { rho }  ]] & := &  [[rho x]] \\
        [[(Pi x : A . B) { rho }]] & := & [[Pi x : A { rho } . B { rho .: x -> x }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ x . a) { rho }]] & := & [[\ x . (a { rho .: x -> x })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

The \lang language is expressive enough to support large
eliminations, the ability to compute a type using a term as input. For
example, the function $[[\ x . if x Bool Void]]$ returns
either $[[Bool]]$ or $[[Void]]$ depending on whether its
input is $[[true]]$ or $[[false]]$.

\yl{Need to fix the de bruijn text}
We adapt from \citet{autosubst2} notation for simultaneous renaming,
substitution, and other auxiliary definitions used in our paper, summarized in
Figure~\ref{fig:auxdef}. We need these definitions to specify the reduction
and typing relations for \lang{}. The metavariable $[[rho]]$ represents
substitutions, which are mappings from de Bruijn indices to terms.  One
example of a substitution is the function $\ottkw{id}_{tm}$, the identity
substitution that maps each index to the corresponding term variable.  % The
% $...$ substitution is more general and adds $[[n]]$ to the index
% before injecting it as a term.
The extension operation, $[[(rho .: x -> a)]]$,
allows us to update the substitution $[[rho]]$ to map the variable
$[[x]]$ to $[[a]]$ rather than $[[rho x]]$.

The substitution operator, which takes the form $[[a {rho}]]$,
traverses the syntax of $[[a]]$ and replaces each variable $[[x]]$
with the term $[[rho x]]$. When traversing under binders (e.g. in the
$[[(\ x . a) { rho }]]$ case), we extend $[[rho]]$ to map the bound
variable $[[x]]$ to itself, protecting it from being replaced.

The substitution operator is referred to as simultaneous substitution as it
substitutes for all variables at once. It is possible to recover single
substitution by composing the extension operator and the identity
substitution: $[[a { b / x }]] := [[a { idtm .: x -> b }]]$.

\yl{Explain why iterated substitution is not good enough}

In our proofs, we find it
more convenient to formulate simultaneous substitution directly rather than
recovering it from single substitution.  In particular, this shows up in the
definition of semantic typing in Section~\ref{sec:logrelproof}, which relies
on simultaneous substitution.

@@

\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 / x } => a1 {
    b1 / x }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b / x } => a1 {b / x}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
% From Lemmas~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
% \ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
% equivalence relation.

@@

\yl{Var, T-Pi, T-Abs are self-explanatory now?}
% In \rref{T-Var}, $G n$ is the partial function defined by the equations
% $ (G ++ A) 0 := A$ and $ (G ++ A) Suc n  := G n$.  The
% precondition $n < | G | $, where $ | G | $ represents the length of
% the context $G$, ensures that $G n$ has a defined output. The type
% $G n$ needs to be shifted by $Suc n$ since the types stored in the
% context $G$ are scoped differently. In the typing judgment
% $G |- a : A$, we want both $a$ and $A$ to have the same
% scope. However, the closer a type is to the head of $G$, the smaller de
% Bruijn it has since there is less variables in the context it can refer to
% (see \rref{Ctx-Cons}). Shifting $G n$ by $Suc n$ weakens the type so
% it is scoped consistently with the full context $G$ rather than a
% truncated version of $G$.

% Although the use of de Bruijn indices complicates the variable rule, the rules
% that involve binding, such as \rref{T-Pi,T-Abs}, are straightforward.

@@

The elimination form for booleans, \rref{T-If}, demonstrates dependent pattern
matching.  The result type of this expression $[[A {a / x}]]$ is composed of some
\emph{motive} $[[A]]$, a type where its single free variable has been replaced
with the condition of the if expression. When typing the true branch, this
substitution replaces the variable by $[[true]]$, and similarly for the false
branch. As a result, the type system communicates the information gained from
the % run-time \yl{commented out run-time per Jonathan's comment}
test to each of the branches of the expression.

@@

@@

A naive attempt to proving Theorem~\ref{theorem:consistency} by
induction on the derivation $[[empty |- a : Void]]$ would succeed at
almost all cases except for \rref{T-App}. In the application
case, we are given $[[empty |- b : Pi x : A . B]]$ and $[[empty |- a : A]]$, and
the equality that $[[B {a / x} = Bool]]$. Our goal is to show that
$[[empty |- b a : Void]]$ is not possible. However, note that there is
nothing we know of $[[b]]$ or $[[a]]$ from the induction hypothesis
because neither $[[Pi x : A . B]]$ nor $[[A]]$ is equal to $[[Bool]]$.
We have no way of deriving a contradiction from $[[empty |- b a :
Void]]$. The takeaway from this failed attempt is that, in order to
derive the consistency, we need to know something about types other
than $[[Void]]$. From a pragmatic point of view, proof by logical
relation can be seen as a sophisticated way of strengthening the
induction hypothesis. From the strengthened property, the fundamental
theorem, we will be able to derive consistency as a corollary.

The complexity of applying proof by logical relation to dependent types stems
from the fact that the logical relation is much harder to define. In
simply typed languages, the logical relation is defined as a recursive
function over the type $[[A]]$.
In dependent types, the type
$[[A]]$ can take the form $[[(\ x . x ) Bool]]$. To assign meaning to
this type, we need to first reduce it to $[[Bool]]$. However, we
cannot write a function that performs the reduction because we do not
know the termination of well-typed terms a priori. As a result, we
define the logical relation as an inductively defined relation,
reminiscent of how we specify the reduction graph of a partial
function; the functionality of the relation can later be recovered in Lemma~\ref{lemma:logreldeter}.

@@

% To derive consistency, it suffices to restrict $[[S]]$ to the set of
% terms that reduce to closed terms.
\Rref{I-Void, I-Bool} capture terms that \emph{behave like} the inhabitants
of the $[[Void]]$ and $[[Bool]]$ types under an empty context. For
example, the $[[Void]]$ type should not have any inhabitants under the
empty context, where as the $[[Bool]]$ type corresponds to terms that
evaluate to $[[true]]$ or $[[false]]$. However, the characterization
of $[[Bool]]$ in our logical relation does not always correspond to
well-typed or even closed terms. For example, we have $[[if false (#
Void true #)
true]]$ is ill-typed under the empty context but still belongs to the set $[[{ a | a =>+
  true \/ a =>+ false }]]$ since it evaluates to $[[true]]$.
The independence of syntactic typing in our logical relation allows
our semantic typing definition in Section~\ref{sec:logrelproof} to be
meaningful on its own. Furthermore, not having to embed scoping
information into the logical relation avoids extra bookkeeping and the
need for a Kripke-style logical relation.
\yl{Not sure what to cite from Derek Dreyer. I know his blog post
  about semantic type soundness but is there a good paper to cite? one
of the rust papers?}

% unsurprising; when considering only closed terms, the empty type
% should not be inhabited and therefore corresponds to the empty set,
% whereas the boolean type is semantically inhabited by terms that
% evaluate to the boolean values $[[true]]$ or $[[false]]$.
\Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresponds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance. \Rref{I-Red} enables us to reduce our types in order
to assign meanings. Recall the type expression $[[(\ x . x )
Bool]]$. \Rref{I-Red} says to know that
$[[Interp I i  (\ x . x) Bool S ]]$ for some $[[S]]$, it suffices to
show that $[[Interp I i Bool S]]$ since $[[(\ x . x) Bool =>
Bool]]$. The derivation that $[[Interp I i (\ x . x) Bool { a | a =>+
  true \/ a =>+ false }]]$ therefore follows by composing \rref{I-Red}
and \rref{I-Bool}.

\Rref{I-Pi} is the most complicated rule in our logical
relation. Instead of explaining it directly, we consider the following
simplified admissible \rref{I-PiAlt} that follows directly from \rref{I-Pi}.
\begin{center}
  \drule[]{I-PiAlt}
\end{center}
\begin{lemma}[I-PiAlt admissibility]
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is admissible.
\end{lemma}
\begin{proof}
Immediate from \rref{I-Pi} with $[[R]]$ instantiated to the relation $[[{ (a
, S0 ) | a in S , Interp I i B { a /x } S0 }]]$.
\end{proof}
Compared to the \rref{I-Pi}, which contains the mysterious relation
$[[R]]$, \rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The precondition of
\rref{I-PiAlt} says that
the function type $[[Pi x : A . B]]$ has an interpretation if its input
type $[[A]]$ can be
interpreted as some set $[[S]]$, and for all terms $[[a in S]]$, the
type $[[B {a / x}]]$, obtained by substituting $[[a]]$ into the output type
$[[B]]$, has some semantic interpretation. In its conclusion, we say
that a term $[[b]]$ belongs to an interpretation of $[[Pi x : A . B]]$ if
for all $[[a in S]]$, where $[[S]]$ is an interpretation $[[A]]$, the
application form $[[b a]]$ belongs to all possible interpretations of
$[[B {a  / x}]]$ (the pre-condition ensures at least one interpretation
exists for each $[[B {a / x}]]$ where $[[a in S]]$).

In fact, while \rref{I-PiAlt} is an instantiation of \rref{I-Pi},
these two rules are equivalent in the sense that every derivation of
involving \rref{I-Pi} can be systematically replaced by
\rref{I-PiAlt}. This equivalence follows directly from the fact
that the logical relation is a partial function, a result we will show
in Lemma~\ref{lemma:logreldeter}. The preconditions of \rref{I-Pi},
when combined with the functionality of the logical relation, uniquely
determine the relation $[[R in S * PowerSet STm]]$ to be the relation
$[[{ (a , S0 ) | a in S , Interp I  i B { a /x } S0 }]]$. This result is
formally shown through the improved inversion lemma for function types
(Lemma~\ref{lemma:piinvalt}).

Unfortunately, we cannot define the function case of our logical
relation directly using \rref{I-PiAlt} since the occurrence of
$[[Interp I i B {a/x} S0]]$ in its conclusion not only violates the
syntactic strict positivity constraint required in proof assistants,
but is genuinely non-monotone when we rewrite the inductive definition
as an endofunction over the domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition as the relation
grows, whereas the
relation $[[R]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ and
prevents the set of witnesses from growing. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties to justify the well-definedness of our
inductive relation with \rref{I-PiAlt}, we opt for our current
formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[R]]$ each time we apply it, though this is mitigated by
the immediate admissibility of \rref{I-PiAlt} and the alternative $\Pi$ inversion
principle (Lemma~\ref{lemma:piinvalt}), which we derive after proving functionality.
